Multi-stage programming with functors and monads: eliminating abstraction overhead from generic code

### 2007-07-13 SCP2008/paper/metamonads.tex
Cut section 2 down to size.  Move some of the text from that section to other places when appropriate. A few minor edits in other sections as well.

In high-performance symbolic and numeric computing, there is a
well-known issue of balancing between maximal performance and the
level of abstraction at which code is written.  Furthermore, already
in linear algebra, there is a wealth of different aspects that
\emph{may} need to be addressed. Implementations of the
widely used LU decomposition algorithm and its particular case Gaussian 
Elimination (GE) \oleg{can I say GE is an instance of LU?} --- the running
example of our paper --- may need to account for the representation of
the matrix, whether to compute and return the determinant or rank, how
and whether search for a pivot, etc. Furthermore, current architectures
demand more and more frequent tweaks which, in general, cannot be done by the
compiler because the tweaking often involves domain knowledge.

@@

The rest of this paper is structured as follows: The next section
gives an overview of the design space of LU decomposition algorithms,
as well as the methodology we follow.  Then section~\ref{CPS}
introduces code generation in MetaOCaml, the problem of name
generation, and continuation-passing style (CPS) as a general
solution.  We also present a key monad and the issues around generating
control statements. Section~\ref{functors} describes the use of
parametrized modules of OCaml to encode all of the aspects of the
LU decomposition algorithm family in completely separate,
independent modules.  We briefly discuss related work in
section~\ref{related}. We then outline the future work and conclude.
Appendices give samples of the generated code (which is available in
full at \cite{metamonadsURL}).

@@

Before investigating implementation approaches, it is worthwhile to
carefully study the design space involved.  A preliminary
study~\cite{Carette06} revealed a number of aspects of the family of
Gaussian Elimination algorithms.  In the present work, we outline a
number of additional aspects involved in the faimily of LU
decomposition algorithms.  These will first be presented in a somewhat
ad hoc manner, roughly coresponding to the order in which they were
``discovered''.  We then reorganize them into groups of
semantically related aspects to form the basis of our design.
Finally, we will extract a methodology from our experience.

@@

  \item \textbf{Input choices}: Whether the input matrix is an augmented
	  matrix or not, and if so, where is the end of the main matrix.
	  \oleg{Is this the same as Augmented matrices above?}

@@

The aspects in the following group have also been observed in
practice but we have not yet implemented them:
\begin{enumerate}
\setcounter{enumi}{\value{naspects}}
\item \textbf{Logging}, a classical cross-cutting concern.
\item \textbf{Sparsity}: If a matrix is known to be sparse, at least
the traversal should be sparse. Maximal preservation of the sparsity
is desirable. 
\item \textbf{Other structure}: If a matrix is known in
advance to be real symmetric tri-diagonal, LU decomposition can be
done in $O(n^2)$ rather than $O(n^3)$ time, at the additional $O(n)$
storage cost.
\item \textbf{Warnings}: If a domain permits only heuristic zero
testing, it is customary to issue a warning (or otherwise log) when 
a potentially zero pivot is chosen.
\item \textbf{In-place}: offering an option of in-place decomposition,
  re-using the input matrix as the storage for the output.
\item \textbf{Error-on-singular}: Raise an exception when the input
  matrix is (near) singular.
\setcounter{naspects}{\value{enumi}}
\end {enumerate}

@@

\omitnow{Clearly the above $\thenaspects$ aspects have a rather complex 
dependency graph.  While some aspects are purely orthogonal to others, 
most are inter-dependent.  For example, if one wants the determinant of
the matrix as part of the output, then the determinant has to be tracked
during the algorithm; a priori, it is possible to make incompatible choices,
and our implementation has to insure that this is in fact impossible.
There are other kinds of dependencies: for example, 
  While there are no circular dependencies, there is
no clear directed graph of dependencies.  In other words, one cannot 
simply use product, co-products, and compositions of the various aspects 
to build the resulting algorithm.  We will get back to this point when
we describe the details of the implementation.}

@@

\oleg{I'm quite uneasy about this and the following sections. It
  appears as `philosophy'. It's OK for us, because it is written in
  the Perfect tense for us. We understand what each of these items
  mean. OTH, readers will read this in the Future tense. They have not
  seen our code and know nothing about our approach. To them, it may
  appear just opaque philosophising.

I'm in favor of drastically abbreviating these sections. Our message
to the reader should be: please keep these groups of aspects as you
read the description of our code below. We also point out that
attempting to separate these aspects and letting the users
independently change them leads -- in the unstaged case -- to such
fantastically inefficient code so that such clean design is rarely
(cite Aldor?) attempted. Staging let us remove the performance
penalties of abstraction and so let us concentrate on the design. 
The last phrase is actually more suitable for conclusion or related
work. I think it would be much better to let related work make these
points for us.}

@@

\begin{enumerate}
	\item \textbf{Abstract domain}.  This group includes
	  `mathematical' aspects: domain of matrix elements, length
	  measure, full division, normalization, zero testing,
	  symmetry and other matrix structure.
	\item \textbf{Concrete representation}: choosing data
	  structures for domains, containers, permutation
	  matrices. This group also includes packing, in-place
	  decomposition, and the
	  representation of sparse matrices.
	\item \textbf{Interface} of each generated function, including
	  the input and output choices, logging and error reporting.
		\oleg{In-place updates is also the issue of concrete
		  representation! As well as of the abstract domains:
		  for some domains we just can't run LU with no extra
		  storage.} 
	\item \textbf{Algorithmic Strategy}, such as fraction-free
	  updates, pivoting strategies, augmented matrices.
	\item \textbf{Tracking}, of determinant, rank, or pivot, etc.
	\item \textbf{Staging}: whether the result if the program or a
	  generator that will produce the program.
\end{enumerate}

\oleg{I'd rather replace this paragraph with my remark above.}
While the first two families of aspects are the most obvious, it is quite
difficult to separate them out cleanly.  This seems to stem from the fact
that in a non-staged setting, separating out these aspects leads to such
fantastically inefficient code, especially in the setting of scientific
computation, that our experience works \emph{against us}.  While it would
be too much to speak of a paradigm shift, it is necessary to firmly think
in terms of abstract ideal design while avoiding all thoughts of traditional
efficiency.

\oleg{I'd rather move this down where we discuss the output choices. I
  want to get top the code quickly. Actually, the following paragraph
  better belongs to related work.}  
Unlike traditional approaches
\cite{journals/cacm/parnas72a}, the interfaces of our generated
routines vary depending on the choices of input and output
aspects. Dynamic approaches in Object-Oriented language (late binding
and dynamic dispatch) or functional languages (Haskell's
dictionary-based \emph{type classes}), too, offer flexibility of
interfaces. In our approach, however, it is the generator that
produces code whose interfaces depend on the arguments of the
generator. The interfaces of the generated routines are all fixed and
hence effecient.

\oleg{I'd drop this: the reader has not seen our implementation.}
The information tracking aspects are in some ways to most intruiging ones:
they are purely internal to the algorithms, and they are the ones which
are most dependent on other choices.  The \emph{implementation} of these
aspects is completely orthogonal to the other aspects, however their
\emph{compositionality properties} are deeply enmeshed with that of other
aspects.  While writing the code for the aspects was straightforward enough,
this is where quite a bit of infrastructure was needed to encode
various dependencies, so that it would be impossible to generate an
instance of the LU Decomposition algorithm with information
tracking aspects which were incompatible with the requirements of other
aspects.

\oleg{I'd move this to the next section.}
Lastly, to ensure that we were indeed firmly in a \emph{cogen} setting,
we abstracted out the underlying programming language completely.  This 
allowed us to both generate (efficient) code and to write a directly
runnable (highly inefficient) version of the algorithm from the same
generator.  This brought out the issue that, once this was done, we had
essentially abstracted out all of the syntactic sugar of the underlying
programming language, and all we were left with was function application
and monadic composition.  This means the code for our generator looks mostly
like Scheme with added syntax for monads!

Our overall approach can best be described a combination 
of two approaches: hand-writing a code generator (cogen) suitable
for multi-level specialization \cite{Gluck95,GluckJ97} and creating
an embedded domain-specific language (EDSL) \cite{edsl,WhenHowDSL}.
\oleg{I'd keep only the above phrase and say that our approach of
  using staging, monads and functors turns out to permit extensible
  set of aspects and flexible enough for iterative improvement in the
  design.}

\oleg{I'm inclined then to drop the following two paragraphs. I'm
  positive reviewers would be upset at philosophical poinst in the
  middle of a technical paper.}
More specifically we start from a \emph{known} set of implementations
of an algorithm, and extract commonalities and variation points.
This is unlike~\cite{journals/tse/Parnas76} and most subsequent
approaches to product families, as we do not over-engineer our design
by imagining weird variations that are unlikely to come up in realistic
situations, but only create variations when we notice them in actual use.
This approach is quite well-suited to the development of scientific
software, which has a rich history and where most useful variations have
already appeared in some form.

@@

The initialization expressions |OD.Det.fin ()| and |IF.R.fin ()| are
the pre-flight checks. As we saw in the previous section, 
both instances of |Det| contain the |fin ()| function to generate the
code representing the computed determinant. Only the |NoDet| does not
track any determinant, and so |fin ()| raises an error. The code 
|let _ = OD.Det.fin ()| in |OutDetRank| invokes that |fin| function, which will
produce the monadic code generating action, or raise the error. We do
not run the action at that time -- we only make sure there is an
action to run. This is the preflight check again, to rule out the
semantic error when the user specified that determinant should be computed
and returned and yet specified the |NoDet| aspect which tracks no
determinant at all.

@@

@@

\section{Conclusion}\label{conclusion}
In this paper we have demonstrated numerical code extensively parameterized
by complex aspects at no run-time overhead.  The combination of
stateless functors and structures, and our monad with the
compositional state makes aspects freely composable without having to
worry about value aliasing. The only constraints to compositionality
are the typing ones plus the constraints we specifically
impose, including semantic constraints (e.g., rings do not have full
division).

@@

\subsection*{Acknowledgments}
We wish to thank Cristiano Calgano for his help in adapting camlp4 for
use with MetaOCaml. Many helpful discussions with Walid Taha are very
appreciated. The implementation of the monadic notation, |perform|,
was the joint work with Lydia van Dijk.

>>>>

In high-performance symbolic and numeric computing, there is a
well-known issue of balancing between maximal performance and the
level of abstraction at which code is written.  Furthermore, already
in linear algebra, there is a wealth of different aspects that
\emph{may} need to be addressed. Implementations of the
widely used LU decomposition algorithm and its particular case Gaussian 
Elimination (GE) --- the running
example of our paper --- may need to account for the representation of
the matrix, whether to compute and return the determinant or rank, how
and whether search for a pivot, etc. Furthermore, current architectures
demand more and more frequent tweaks which, in general, cannot be done by the
compiler because the tweaking often involves domain knowledge.

@@

The rest of this paper is structured as follows: The next section
gives an overview of the design space of LU decomposition algorithms.
Then section~\ref{CPS}
introduces code generation in MetaOCaml, the problem of name
generation, and continuation-passing style (CPS) as a general
solution.  We also present a key monad and the issues around generating
control statements. Section~\ref{functors} describes the use of
parametrized modules of OCaml to encode all of the aspects of the
LU decomposition algorithm family in completely separate,
independent modules.  We briefly discuss related work in
section~\ref{related}. We then outline future work. In our
conclusion (section~\ref{conclusion}), we sum up our work, comment
on programmings with aspects, and outline the overall methodology
we used to structure our work.
Appendices give samples of the generated code (which is available in
full at \cite{metamonadsURL}).

@@

Before investigating implementation approaches, it is worthwhile to
carefully study the design space involved.  A preliminary
study~\cite{Carette06} revealed a number of aspects of the family of
Gaussian Elimination algorithms.  In the present work, we outline a
number of additional aspects involved in the faimily of LU
decomposition algorithms.  These will first be presented in a somewhat
ad hoc manner, roughly coresponding to the order in which they were
``discovered''.  We then reorganize them into groups of
semantically related aspects to form the basis of our design.

@@

  \item \textbf{Input choices}: Grouping all the potential
	  choices of inputs -- currently only augmented matrices require
	  an extra input.

@@

The aspects in the following group have also been observed in
practice but we have not yet implemented them:
\begin{enumerate}
\setcounter{enumi}{\value{naspects}}
\item \textbf{Logging}, a classical cross-cutting concern.
\item \textbf{Sparsity}: If a matrix is known to be sparse, at least
the traversal should be sparse. Maximal preservation of the sparsity
is desirable. 
\item \textbf{Other structure}: If a matrix is known in
advance to be real symmetric tri-diagonal, LU decomposition can be
done in $O(n^2)$ rather than $O(n^3)$ time, at an additional $O(n)$
storage cost.
\item \textbf{Warnings}: If a domain permits only heuristic zero
testing, it is customary to issue a warning (or otherwise log) when 
a potentially zero pivot is chosen.
\item \textbf{In-place}: offering an option of in-place decomposition,
  re-using the input matrix as the storage for the output.
\item \textbf{Error-on-singular}: Raise an exception when the input
  matrix is (near) singular.
\setcounter{naspects}{\value{enumi}}
\end {enumerate}

@@

@@

@@

\begin{enumerate}
	\item \textbf{Abstract domain}.  This group includes
	  `mathematical' aspects: domain of matrix elements, length
	  measure, full division, normalization, zero testing,
	  symmetry and other matrix structure.
	\item \textbf{Concrete representation}: choosing data
	  structures for domains, containers, permutation
	  matrices. This group also includes packing, in-place
	  decomposition, and the
	  representation of sparse matrices.
	\item \textbf{Interface} of each generated function, including
	  the input and output choices, logging and error reporting.
	\item \textbf{Algorithmic Strategy}, such as fraction-free
	  updates, pivoting strategies, augmented matrices.
	\item \textbf{Tracking}, of determinant, rank, or pivot, etc.
	\item \textbf{Staging}: whether the result if the program or a
	  generator that will produce the program.
\end{enumerate}

\oleg{In-place updates is also the issue of concrete
  representation! As well as of the abstract domains:
  for some domains we just can't run LU with no extra
  storage.} 
\jacques{Correct.  There are issues which are ``relational'' rather
than being completely independent.  The above categories seem to 
be independent, but that is too simplistic.  Some aspects, like
in-place, have quite a number of pre-conditions on other aspects
before they can be realized.}

@@

The initialization expressions |OD.Det.fin ()| and |IF.R.fin ()| are
the pre-flight checks. As we saw in the previous section, 
both instances of |Det| contain the |fin ()| function to generate the
code representing the computed determinant. Only the |NoDet| does not
track any determinant, and so |fin ()| raises an error. The code 
|let _ = OD.Det.fin ()| in |OutDetRank| invokes that |fin| function, which will
produce the monadic code generating action, or raise an error. We do
not run the action at that time -- we only make sure there is an
action to run. This is another preflight check, to rule out the
semantic error when a user specifies that the determinant should be computed
and returned and yet specifies the |NoDet| aspect which tracks no
determinant at all.

It is also worth noticing the |INTERNAL_FEATURES| type, which bundles
various information tracking aspects.  These are in some ways rather
intruiging: they are purely internal to the algorithms, yet they are the
ones which
are most dependent on other choices.  The \emph{implementation} of these
aspects is completely orthogonal to the other aspects, however their
\emph{compositionality properties} are deeply enmeshed with that of other
aspects.  While writing the code for these was straightforward enough,
this required quite a bit of type-level infrastructure, as well as the
preflight checks%
\footnote{We had originally~\cite{CaretteKiselyov05} implemented all our
preflight tests at the type level, but this did not scale.}
to get everything to work properly.

@@

Unlike traditional approaches
\cite{journals/cacm/parnas72a}, the interfaces of our generated
routines vary depending on the choices of input and output
aspects. Dynamic approaches in Object-Oriented language (late binding
and dynamic dispatch) or functional languages (Haskell's
dictionary-based \emph{type classes}) also offer flexibility of
interfaces. In our approach, however, it is the generator that
produces code whose interfaces depend on the arguments of the
generator. The interfaces of the generated routines are all fixed and
hence efficient.

@@

\section{Conclusion}\label{conclusion}
In this paper we have demonstrated numerical code extensively parameterized
by complex aspects at no run-time overhead.  The combination of
stateless functors and structures, and our monad with the
compositional state makes aspects composable without having to
worry about value aliasing. The only constraints to compositionality
are the typing ones plus the constraints we specifically
impose, including semantic constraints (e.g., rings do not have full
division).

\subsection{On aspects}

@@

While the first two families of aspects (abstract domain and concrete
representation) are the most obvious, it is quite
difficult to separate them out cleanly.  This seems to stem from the fact
that in a non-staged setting, separating out these aspects leads to such
fantastically inefficient code, especially in the setting of scientific
computation, that our experience works \emph{against us}.  While it would
be too much to speak of a paradigm shift, it is necessary to firmly think
in terms of abstract ideal design while avoiding all thoughts of traditional
efficiency.  Interestingly, in conversations that the first author had
with D. Parnas, we discovered this is apparently
what~\cite{journals/cacm/parnas72a} was advocating.  We believe that we are
taking the first steps towards a \emph{typed yet efficient} realization of
these ideas, where the various design-time entities can be directly encoded in
machine-checkable form.

Our overall approach can best be described a combination 
of two approaches: hand-writing a code generator (cogen) suitable
for multi-level specialization \cite{Gluck95,GluckJ97} and creating
an embedded domain-specific language (EDSL) \cite{edsl,WhenHowDSL}.
Our approach, using staging, monads and functors, seems to permit 
an extensible set of aspects and appears flexible enough for iterative
improvement in the design.

As we wanted to ensure that we were indeed firmly in a \emph{cogen} setting,
we abstracted out the underlying programming language completely.  This 
allowed us to both generate (efficient) code and to write a directly
runnable (highly inefficient) version of the algorithm from the same
generator.  This brought out the issue that, once this was done, we had
essentially abstracted out all of the syntactic sugar of the underlying
programming language, and all we were left with was function application
and monadic composition.  This means the code for our generator looks mostly
like Scheme with added syntax for monads!

More specifically we start from a \emph{known} set of implementations
of an algorithm, and extract commonalities and variation points.
This is unlike~\cite{journals/tse/Parnas76} and most subsequent
approaches to product families, as we do not over-engineer our design
by imagining weird variations that are unlikely to come up in realistic
situations, but only create variations when we notice them in actual use.
This approach is quite well-suited to the development of scientific
software, which has a rich history and where most useful variations have
already appeared in some form.

\subsection*{Acknowledgments}
We wish to thank Cristiano Calgano for his help in adapting camlp4 for
use with MetaOCaml. Many helpful discussions with Walid Taha are very
appreciated. The implementation of the monadic notation, |perform|,
was joint work with Lydia van Dijk.
